---
title: "Zero-step timer analysis — inducible promoter characterisation"
author: "Rebecca Wolters"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

# Setup

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(MetBrewer)
library(wellr)       # github: bradyajohnston/wellr
library(broom)
library(scales)
library(reluxr)
library(wesanderson)
library(patchwork)
library(here)
```

## Load functions

```{r source}
source(here("zerostep-timer", "zerosteptimer_functions_clean.R"))
```

## Experiment parameters

All experiment-level parameters are defined here. Change only this chunk
when analysing a new experiment.

The zero-step timer measures inducible promoters (Ptet, Pbad, PpcaU) driving
luciferase directly. Key metadata columns are:
- `promoter`      — inducible promoter (e.g. "Ptet", "Pbad", "PpcaU")
- `copy`          — plasmid copy number ("sc" = single copy, "mc" = multi copy)
- `concentration` — inducer concentration (units depend on promoter)
- `strain`        — strain ID

```{r parameters}
# ---- Timing ----------------------------------------------------------------
it        <- 2      # induction time (hours)
starttime <- 0      # time considered t=0

# ---- Analysis --------------------------------------------------------------
fc_threshold <- 4   # fold-change threshold for time delay calculation

# ---- Plate format ----------------------------------------------------------
# Pathlength correction factors (determined empirically)
# 50 µl in 384-well = 3.09 | 100 µl in 96-well = 4.898
pc_factor <- 4.898  # 96-well plates used for zero-step timer

# ---- File paths ------------------------------------------------------------
data_files <- here("zerostep-timer", "data")
plot_dir   <- here("zerostep-timer", "plots")
# ---- Grouping variables ----------------------------------------------------
# Note: zero-step uses promoter/copy/concentration instead of
# antisigma/arabinose/aTc/DHBA as seen in the timer analysis
group_vars_av <- c("time", "strain", "copy", "promoter", "concentration")
group_vars    <- c("strain", "copy", "promoter", "concentration")

# ---- Plot axis breaks -------------------------------------------------------
breaks       <- 10^(-10:10)
minor_breaks <- rep(1:9, 21) * (10^rep(-10:10, each = 9))
```

---

# Data loading

## Raw plate reader data

```{r load-raw}
df_raw <- read_in_plates(data_files, pattern = "RW.csv")
df_raw
```

## Metadata

```{r load-meta}
mdf <- read_in_meta(data_files, pattern = "metainfo.csv")
mdf
```

## Join raw data and metadata

```{r join}
df <- df_raw |>
  dplyr::left_join(mdf, by = c("well" = "Wells", "exp" = "exp")) |>
  na.omit()
df
```

## Pull strain names from metainfo

```{r strain-names}
strains <- df |>
  dplyr::filter(!strain %in% c("blank"), !is.na(strain)) |>
  dplyr::pull(strain) |>
  unique() |>
  sort()

strains
```

---

# Blanking

## Apply blank correction

```{r blank-correct}
# Update average_blank_od from the get_blank_average() output above
# 96-well empirical blank OD: 0.034
df_blank <- blank_plate(
  data             = df,
  average_blank_od = 0.034,
  pc_factor        = pc_factor
)

df_blank
```

## QC: Plate overview plots

```{r plate-overview-od, fig.width=14, fig.height=10}
# Change number_of_experiment to check different plate runs
plot_plate(df_blank,
           variable_to_plot     = "od600_blank",
           variable_to_color    = "promoter",
           inductiontime        = it,
           number_of_colors     = 3,
           number_of_experiment = 2)
```

```{r plate-overview-lum, fig.width=14, fig.height=10}
plot_plate(df_blank,
           variable_to_plot     = "lum_blank",
           variable_to_color    = "promoter",
           inductiontime        = it,
           number_of_colors     = 3,
           number_of_experiment = 2)
```

---

# Averaging

```{r average}
df_av <- average_plate(df_blank, group_vars = group_vars_av)
df_av
```

## Check replicate counts

```{r replicate-counts}
df_av |>
  dplyr::mutate(count = stringr::str_count(replicate, pattern = "[A-Z][0-9]{2}")) |>
  dplyr::summarise(av_count = mean(count), .by = strain)
```

---

# Background luminescence

```{r background}
bg_expr <- df_av |>
  get_background_lum(
    value         = lum_blank_mean,
    starttime     = starttime,
    inductiontime = it,
    group_vars    = group_vars,
    wildtype      = FALSE
  )

bg_expr
```

---

# Averaged kinetics plots

## OD600

```{r plot-od, fig.width=7, fig.height=4}
# Filter to strains of interest before calling — example shows Ptet strains
df_av |>
  dplyr::filter(strain %in% c("GFC0200", "GFC0310")) |>
  plot_average_od(
    variable_to_colour = "concentration",
    unit_of_colour     = "ng/ml",
    strain_to_plot     = "Ptet",
    inductiontime      = it,
    grid_col           = "copy",
    grid_row           = "promoter"
  )
```

## Luminescence kinetics

```{r plot-lum, fig.width=7, fig.height=4}
df_av |>
  dplyr::filter(strain %in% c("GFC0200", "GFC0310")) |>
  plot_average_lum(
    variable_to_colour = "concentration",
    unit_of_colour     = "ng/ml",
    inductiontime      = it,
    background_lum     = 85,
    grid_col           = "copy",
    grid_row           = "promoter"
  )
```

---

# Growth rate

```{r growthrate}
df_gr <- get_growthrate(
  df_blank,
  group_vars = c("strain", "promoter", "copy", "concentration", "exp"),
  starttime  = 0,
  endtime    = 2
)

df_gr_av <- average_gr_data(
  df_gr,
  group_vars = c("strain", "promoter", "copy", "concentration")
)

df_gr_av
```

---

# Time delays

## Calculate interpolated time delays (recommended)

Uses cubic spline interpolation to find the precise time at which
luminescence crosses `fc_threshold` × pre-induction baseline.

```{r timedelay-spline}
df_timedelay_all <- df_av |>
  dplyr::filter(strain != "blank") |>
  get_timedelay_splinefun(
    bg_expr       = bg_expr,
    group_vars    = group_vars,
    foldchange    = fc_threshold,
    inductiontime = it
  )

df_timedelay_all
```

## Plot luminescence kinetics with interpolated time delay

```{r plot-lum-timedelay, fig.width=9, fig.height=5}
# Example: PpcaU strains
strain_of_interest <- c("GFC0153", "GFC0509") 
  
  #c("GFC0698", "GFC0699") - PpcaU strains
  #c("GFC0200","GFC0310") - Ptet strains
  #c("GFC0153", "GFC0509") - Pbad strains

plot_lumkinetic_with_timedelay(
  .data              = df_av,
  strain_to_plot     = strain_of_interest,
  df_timedelay       = df_timedelay_all,
  inductiontime      = it,
  variable_to_colour = "concentration",
  unit_of_colour     = "%",
  background_lum     = 85,
  grid_col           = "copy",
  grid_row           = "promoter",
  group_vars         = group_vars
) +
  ggplot2::annotation_logticks(sides = "l")

ggplot2::ggsave(
  filename = file.path(plot_dir, "Pbad_LUM_kinetic_timedelay_interpolated.pdf"),
  width = 80, height = 35, units = "mm", dpi = 300, scale = 3
)
```

## Plot all strains

```{r plot-lum-all-strains, message=FALSE}
# Loop through all strains and save individual plots 
for (strain_name in strains) {

  plt <- plot_lumkinetic_with_timedelay(
    .data              = df_av,
    strain_to_plot     = strain_name,
    df_timedelay       = df_timedelay_all,
    inductiontime      = it,
    variable_to_colour = "concentration",
    unit_of_colour     = "µM",
    background_lum     = 85,
    grid_col           = "copy",
    grid_row           = "promoter",
    group_vars         = group_vars
  ) +
    ggplot2::annotation_logticks(sides = "l")

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("0steptimer_", strain_name, "_LUM_interpolated.pdf")),
    plot     = plt,
    width = 80, height = 35, units = "mm", dpi = 300, scale = 3
  )
}
```

---

# Dynamic range

## Calculate fold-change at all time points

```{r dynamic-range-all}
df_av_fc_all <- get_dynamic_range_all(df_av, bg_expr, group_vars)
```

## Join time delays and fold-changes at endpoint

```{r delays-fc}
# Filter to final time point before joining to avoid row fan-out
df_fc_endpoint <- df_av_fc_all |>
  dplyr::filter(
    time == max(time),
    .by  = all_of(group_vars)
  )

df_delays_fc <- df_timedelay_all |>
  dplyr::left_join(df_fc_endpoint, by = group_vars)

# Sanity check: row counts should match
nrow(df_timedelay_all)
nrow(df_delays_fc)

df_delays_fc
```

## Maximum dynamic range bar plots

```{r dynamic-range-bar}
df_av_fc <- average_dynamic_range(df_blank, bg_expr, group_vars)

# NEED TO CHANGE the promoter filter in the plot function input and the plot name manually! 

# Example: filter to Pbad promoter only
plot_dr <- df_av_fc |>
  plot_dynamic_range_averages(promoter_filter = "PpcaU") # change depending on what promoter to look at

plot_dr

ggplot2::ggsave(
  filename = file.path(plot_dir, "0steptimer_PpcaU_dynamicrange_barplot.pdf"), #change depending on what promoter to look at
  plot     = plot_dr,
  width = 180, height = 120, units = "mm", dpi = 300
)
```

---

# Delay vs fold-change scatter plot

```{r delay-vs-fc, fig.width=5, fig.height=4}
df_delays_fc |>
  plot_delay_vs_fc(inductiontime = it, legend = FALSE) +
  ggplot2::annotation_logticks(sides = "l")

ggplot2::ggsave(
  filename = file.path(plot_dir, "0steptimer_foldchange_vs_timedelay.pdf"),
  width = 80, height = 50, units = "mm", dpi = 300
)
```

---

# Dose-response curves

```{r dose-response-ptet, fig.width=5, fig.height=3}
# Ptet — induced with aTc (ng/ml)
plot_dr_ptet <- df_av |>
  plot_doseresponse(
    promoter_name   = "Ptet",
    inducer         = "aTc",
    unit_of_inducer = "ng/ml"
  )

plot_dr_ptet

ggplot2::ggsave(
  filename = file.path(plot_dir, "0steptimer_Ptet_doseresponse.pdf"),
  plot     = plot_dr_ptet,
  width = 80, height = 45, units = "mm", dpi = 300, scale = 1.5
)
```

```{r dose-response-pbad, fig.width=5, fig.height=3}
# Pbad — induced with arabinose (%)
plot_dr_pbad <- df_av |>
  plot_doseresponse(
    promoter_name   = "Pbad",
    inducer         = "arabinose",
    unit_of_inducer = "%"
  )

plot_dr_pbad

ggplot2::ggsave(
  filename = file.path(plot_dir, "0steptimer_Pbad_doseresponse.pdf"),
  plot     = plot_dr_pbad,
  width = 80, height = 45, units = "mm", dpi = 300, scale = 1.5
)
```

```{r dose-response-ppcau, fig.width=5, fig.height=3}
# PpcaU — induced with DHBA (µM)
plot_dr_ppcau <- df_av |>
  plot_doseresponse(
    promoter_name   = "PpcaU",
    inducer         = "DHBA",
    unit_of_inducer = "µM"
  )

plot_dr_ppcau

ggplot2::ggsave(
  filename = file.path(plot_dir, "0steptimer_PpcaU_doseresponse.pdf"),
  plot     = plot_dr_ppcau,
  width = 80, height = 45, units = "mm", dpi = 300, scale = 1.5
)
```

## Export dose-response data

```{r dose-response-export}
df_doseresponse <- df_av |>
  dplyr::filter(time == max(time)) |>
  dplyr::select(strain, copy, promoter, concentration, lum_blank_mean, lum_blank_sd)

readr::write_csv(
  df_doseresponse,
  file.path(plot_dir, "0step_doseresponse.csv")
)
```
