---
title: "Single-step timer analysis"
author: "Rebecca Wolters"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

# Setup

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(MetBrewer)
library(wellr)       # github: bradyajohnston/wellr
library(broom)
library(scales)
library(reluxr)
library(wesanderson)
library(ggpubr)
library(cowplot)
library(patchwork)
```

## Load functions

```{r source}
source(here("singlestep-timer", "singlesteptimer_functions_clean.R"))
```

## Experiment parameters

All experiment-level parameters are defined here. Change only this chunk
when analysing a new experiment.

```{r parameters}
# ---- Timing ----------------------------------------------------------------
it         <- 2      # induction time (hours)
starttime  <- 0      # time considered as t=0 (0 for 96-well, 4 for 384-well)

# ---- Analysis --------------------------------------------------------------
fc_threshold <- 4    # fold-change threshold for time delay calculation

# ---- Plate format ----------------------------------------------------------
# Pathlength correction factors (determined empirically)
# 50 µl in 384-well = 3.09 | 100 µl in 96-well = 4.898
pc_factor <- 4.898   # change to pc_factor_384well = 3.09 for 384-well data

# ---- File paths ------------------------------------------------------------
# ---- Change depending on what data-set to analyse (reporter on chromosome - Stefano, or reporter on plasmid - Rebecca)
data_files <- here("singlestep-timer", "data_Rebecca/96-well") # change to 96-well vs 384-well for Rebecca's data 
plot_dir   <- here("singlestep-timer", "plots_Rebecca/96-well")
# ---- Grouping variables ----------------------------------------------------
# group_vars_av must include "time" (used for averaging)
# group_vars is used for all downstream joins and summaries
# Add "DHBA" to both vectors when analysing PpcaU/DHBA data
group_vars_av <- c("time", "strain", "antisigma", "arabinose", "aTc", "DHBA")
group_vars    <- c("strain", "antisigma", "arabinose", "aTc", "DHBA")

# ---- Strain groups ---------------------------------------------------------
# manually defining the strains that we are interested in
# later in the script we will pull out all strain names out of the actually used dataframe as well
# Ptet circuits are induced with aTc; PpcaU circuits with DHBA 
# SV = Stefano's strains; RW = Rebecca's strains
#
Ptet_strains  <- c("GFC0706", "GFC0707", "GFC0708", "GFC0736")
PpcaU_strains <- c("GFC0709", "GFC0710" , "GFC0711", "GFC0737") #

# ---- Plot axis breaks -------------------------------------------------------
breaks        <- 10^(-10:10)
minor_breaks  <- rep(1:9, 21) * (10^rep(-10:10, each = 9))
```

---

# Data loading

## Raw plate reader data

```{r load-raw}
df_raw <- read_in_plates(data_files, pattern = "RW.csv")
# Use pattern = "RW.csv" for Rebecca's data
df_raw
```

## Metadata

```{r load-meta}
mdf <- read_in_meta(data_files, pattern = "metainfo.csv")
mdf
```

## Join raw data and metadata

```{r join}
df <- df_raw |>
  dplyr::left_join(mdf, by = c("well" = "Wells", "exp" = "exp")) |>
  na.omit()
df
```
## Pull out Stefano's strain names for plotting purposes

```{r}
# Pull all unique non-control strain names directly from the data
# This automatically captures all of Stefano's strains without manual listing
strains <- df |>
  dplyr::filter(
    !strain %in% c("blank", "SV01"),   # exclude controls
    !is.na(strain)
  ) |>
  dplyr::pull(strain) |>
  unique() |>
  sort()                                # sort alphabetically for reproducibility

strains
```


---

# Blanking

## QC: Check blank readings before induction (done very initially with Stefano's measurments, blanks were stable and not included in following plate reader assays)

```{r check-blank}
check_blank(df,
            blank_variable    = "lum",
            preinduction_time = it,
            variable_to_color = "exp")
```

## Get average blank OD value

```{r blank-average}
get_blank_average(df,
                  preinduction_time  = it,
                  blank_lum_or_od600 = od600)
```

## Apply blank correction

Use `blank_plate()` when no blank well is on the plate.
Use `blank_on_plate()` when a blank well was included on every plate.

```{r blank-correct}
# No blank well on plate — supply empirical blank OD
# 96-well empirical blank OD: 0.034 | 384-well path-corrected: 0.118
df_blank <- blank_plate(
  data             = df,
  average_blank_od = 0.034,   # update from get_blank_average() output above
  pc_factor        = pc_factor
)

df_blank
```

## QC: Plate overview plots

```{r plate-overview-od, fig.width=12, fig.height=8}
# Change number_of_experiment to check different plate runs
plot_plate(df_blank,
           variable_to_plot    = "od600_blank",
           variable_to_color   = "strain",
           inductiontime       = it,
           number_of_colors    = 12,
           number_of_experiment = 2)
```

```{r plate-overview-lum, fig.width=12, fig.height=8}
plot_plate(df_blank,
           variable_to_plot    = "lum_blank",
           variable_to_color   = "strain",
           inductiontime       = it,
           number_of_colors    = 12,
           number_of_experiment = 4)
```

## QC: Check replicates for a single strain

```{r check-replicates}
df_blank |>
  dplyr::filter(strain == "GFC0706") |>
  plot_replicates_lum(
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    inductiontime      = it,
    background_lum     = 80.8,
    grid_col           = interaction(strain, exp),
    grid_row           = rep
  )
```

---

# Averaging

```{r average}
df_av <- average_plate(df_blank, group_vars = group_vars_av)
df_av
```

## Check replicate counts

```{r replicate-counts}
df_av |>
  dplyr::mutate(count = stringr::str_count(replicate, pattern = "[A-Z][0-9]{2}")) |>
  dplyr::summarise(av_count = mean(count), .by = strain)
```

---

# Background luminescence

Background is calculated from the pre-induction window (starttime to it)
using each strain's own signal (`wildtype = FALSE`).

```{r background}
bg_expr <- df_av |>
  get_background_lum(
    value         = lum_blank_mean,
    starttime     = starttime,
    inductiontime = it,
    group_vars    = group_vars,
    wildtype      = FALSE
  )

bg_expr
```

---

# Averaged kinetics plots

## OD600

```{r plot-od, message=FALSE}
strains <- df_av |>
  dplyr::filter(strain != "blank") |>
  dplyr::pull(strain) |>
  unique()

for (strain_name in Ptet_strains) {
  
  plt <- plot_average_od(
    df_av,
    strain_to_plot     = strain_name,
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    inductiontime      = it,
    grid_col           = aTc,
    grid_row           = antisigma
  )
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_OD.pdf")),
    plot     = plt,
    width    = 180, height = 100, units = "mm", dpi = 300, scale = 1.5
  )
}

# Specs for plots 
# For 96-well assays 
# width    = 80, height = 50, units = "mm", dpi = 300, scale = 1.5

# For 384-well assays
#width    = 180, height = 50, units = "mm", dpi = 300, scale = 1.5
```

## Luminescence (overview — all strains)

```{r plot-lum-overview, fig.width=10, fig.height=8}
df_av |>
  dplyr::filter(strain != "blank") |>
  plot_average_lum(
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    inductiontime      = it,
    background_lum     = 82.8,
    grid_col           = aTc,
    grid_row           = antisigma
  )
```

---

# Growth rate

```{r growthrate}
df_gr <- get_growthrate(
  df_blank,
  group_vars = c("strain", "antisigma", "aTc", "DHBA", "arabinose", "rep", "exp"),
  starttime  = 4,
  endtime    = 7
)

df_gr_av <- average_gr_data(df_gr, group_vars = group_vars)

# Check wildtype growth rate
df_gr_av |> dplyr::filter(strain == "SV01")
```

---

# Time delays

## Calculate interpolated time delays (recommended)

Uses cubic spline interpolation to find the precise time at which
luminescence crosses `fc_threshold` × pre-induction baseline.
This removes quantisation error from discrete measurement intervals.

```{r timedelay-spline}
df_timedelay_all <- df_av |>
  dplyr::filter(strain != "blank") |>
  get_timedelay_splinefun(
    bg_expr       = bg_expr,
    group_vars    = group_vars,
    foldchange    = fc_threshold,
    inductiontime = it
  )

df_timedelay_all
```

## Plot luminescence kinetics with interpolated time delay

```{r plot-lum-timedelay, message=FALSE}
# Change Ptet_strains to PpcaU_strains or strains as needed
for (strain_name in Ptet_strains) {
  
  plt <- plot_lumkinetic_with_timedelay(
    .data              = df_av,
    strain_to_plot     = strain_name,
    df_timedelay       = df_timedelay_all,
    inductiontime      = it,
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    background_lum     = 82.8,
    grid_col           = "aTc",
    grid_row           = "antisigma",
    group_vars         = group_vars
  ) +
    ggplot2::annotation_logticks(sides = "l")
  
  ggplot2::ggsave(
    filename = file.path(
      plot_dir,
      paste0("1steptimer_", strain_name, "_LUM_interpolated.pdf")
    ),
    plot   = plt,
    width  = 180, height = 100, units = "mm", dpi = 300, scale = 1.5
  )
}

# Specs for the plots 
# Use this for the 96-well assays 
# width  = 100, height = 60, units = "mm", dpi = 300, scale = 1.5

# Use this for the 

```


## Plot lum with time delay and additional with and without aTc overlay 

```{r}

for (strain_name in PpcaU_strains) { # change depending on what strains to look at in Rebecca's data
  
  p_kinetics <- plot_lumkinetic_with_timedelay(
    .data              = df_av,
    strain_to_plot     = strain_name,
    df_timedelay       = df_timedelay_all,
    inductiontime      = it,
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    background_lum     = 82.8,
    grid_col           = "DHBA", #change to 'aTc' or 'DHBA' depending on which strains to plot
    grid_row           = "antisigma",
    group_vars         = group_vars
  )
  
  p_overlay <- plot_lumkinetic_atc_overlay(
    .data             = df_av,
    strain_to_plot    = strain_name,
    df_timedelay      = df_timedelay_all,
    inductiontime     = it,
    background_lum    = 82.8,
    group_vars        = group_vars,
    threshold_inducer = "DHBA", #change to 'aTc' or 'DHBA' depending on which strains to plot
    inducer_low       = 0,
    inducer_high      = 2.5, #change to '2.5' for aTc and '250'for DHBA depending on which strains to plot
    grid_row          = "antisigma"
  )
  
  plt <- p_kinetics + p_overlay +
    patchwork::plot_layout(widths = c(2, 1)) 
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_LUM_reviewer_comparison.pdf")),
    plot = plt, width = 180, height = 80, units = "mm", dpi = 300
  )
}

```




## Plot time delay heatmaps

```{r plot-delay-heatmap, message=FALSE}
for (strain_name in PpcaU_strains) {  # change depending on what strains to look at in Rebecca's data
  
  plt <- df_timedelay_all |>
    dplyr::filter(strain == strain_name) |>
    plot_timedelay_heatmap(
      fill_variable = delay,
      x_data        = "DHBA",      # change to "DHBA" for PpcaU strains, 'aTc' for Ptet strains
      y_data        = "arabinose",
      x_unit        = "µM", # change unit for aTc to 'ng/ml' and 'µM' for DHBA
      y_unit        = "%",
      inductiontime = it
    )
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_delay_heatmap.pdf")),
    plot     = plt,
    width    = 80, height = 40, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Overlay plots: with vs without antisigma (max arabinose)

```{r plot-overlay, message=FALSE}
# Uses non-interpolated delays for the overlay — switch to df_timedelay_all if preferred
for (strain_name in PpcaU_strains) { # change depending on what strains to look at in Rebecca's data
  
  df_timedelay_maxarab <- df_av |>
    dplyr::filter(arabinose == max(arabinose), strain == strain_name) |>
    get_timedelay_splinefun(
      bg_expr       = bg_expr,
      group_vars    = group_vars,
      foldchange    = fc_threshold,
      inductiontime = it
    )
  
  plt <- df_av |>
    plot_lum_with_withoutAS_overlay(
      strain_to_plot = strain_name,
      df_timedelay   = df_timedelay_maxarab,
      inductiontime  = it,
      background_lum = 82.5, 
      variable_to_colour = "DHBA" , 
      unit_of_colour = "µM"
    ) +
    ggplot2::annotation_logticks(sides = "l")
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_LUM_plusminusAS.pdf")),
    plot     = plt,
    width    = 180, height = 50, units = "mm", dpi = 300, scale = 1.5
  )
}
```

---

# Dynamic range

## Calculate fold-change at all time points

```{r dynamic-range-all}
df_av_fc_all <- get_dynamic_range_all(df_av, bg_expr, group_vars)
df_av_fc_all
```

## Join time delays and fold-changes

```{r delays-fc}
# Maximum fold-change at final time point per condition
df_fc_endpoint <- df_av_fc_all |>
  dplyr::filter(
    time == max(time),
    .by = all_of(group_vars)
  )

# Now join — one row per group in both tables
df_delays_fc <- df_timedelay_all |>
  dplyr::left_join(df_fc_endpoint, by = group_vars)

df_delays_fc
```

## Dynamic range heatmaps

```{r dynamic-range-heatmap, message=FALSE}
# Change PpcaU_strains to Ptet_strains or strains as needed
# Change x_data to "aTc" and x_unit to "ng/ml" for Ptet strains
for (strain_name in PpcaU_strains) {
  
  plt <- df_av_fc_all |>
    dplyr::summarise(
      max_dynamicrange = max(fc_mean),
      .by              = all_of(group_vars)
    ) |>
    dplyr::filter(strain == strain_name) |>
    plot_dynamic_range_heatmap(
      fill_variable = max_dynamicrange,
      x_data        = "DHBA", # change to 'aTc' or 'DHBA'
      y_data        = "arabinose",
      x_unit        = "µM", # change unit for aTc to 'ng/ml' and 'µM' for DHBA
      y_unit        = "%"
    )
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_foldchange_heatmap.pdf")),
    plot     = plt,
    width    = 80, height = 40, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Maximum dynamic range bar plots

```{r}
# Define your desired left-to-right order of the barplot 
antisigma_order_Ptet <- c("Ptet-AS20_t1", "Ptet-AS22_t", "Ptet-AS28_t1", "Ptet-AS31_t1")  
antisigma_order_PpcaU <- c("PpcaU-AS20_t1", "PpcaU-AS22_t", "PpcaU-AS28_t1", "PpcaU-As31_t1")
```


```{r dynamic-range-bar}
df_av_fc <- average_dynamic_range(df_blank, bg_expr, group_vars)

# Ptet strains (aTc threshold)
plot_Ptet <- df_av_fc |>
  dplyr::filter(!antisigma %in% c("blank", "Ptet-lux", "Pbad-lux"),
                strain     != "SV01") |>
  dplyr::mutate(antisigma = factor(antisigma, levels = antisigma_order_Ptet)) |>
  plot_dynamic_range_averages(
    strain_group      = Ptet_strains,
    threshold_inducer = "aTc"
  )

plot_Ptet

ggplot2::ggsave(
  filename = file.path(plot_dir, "1steptimer_dynamicrange_Ptet-AS_barplot.pdf"),
  plot     = plot_Ptet,
  width = 90, height = 45, scale= 2,units = "mm", dpi = 300
)

# PpcaU strains (DHBA threshold)
plot_PpcaU <- df_av_fc |>
  dplyr::filter(!antisigma %in% c("blank", "Ptet-lux", "Pbad-lux"),
                strain     != "SV01") |>
  dplyr::mutate(antisigma = factor(antisigma, levels = antisigma_order_PpcaU)) |>
  plot_dynamic_range_averages(
    strain_group      = PpcaU_strains,
    threshold_inducer = "DHBA"
  )

plot_PpcaU

ggplot2::ggsave(
  filename = file.path(plot_dir, "1steptimer_dynamicrange_PpcaU-AS_barplot.pdf"),
  plot     = plot_PpcaU,
  width = 90, height = 45, scale = 2, units = "mm", dpi = 300
)
```
## Maximum dynamic range bar plots with annotation 
```{r}
plot_annotation <- plot_dynamic_range_with_annotation(
  .data             = df_av_fc,
  strain_group      = strains,
  threshold_inducer = "aTc",
  sig_results       = df_atc_sign
)

ggplot2::ggsave(
  filename = file.path(plot_dir, "1steptimer_dynamicrange_Ptet-AS_barplot_with_annotation.pdf"),
  plot     = plot_annotation,
  width = 180, height = 80, scale = 2, units = "mm", dpi = 300)
```


---

# Delay vs fold-change scatter plot

```{r delay-vs-fc, fig.width=5, fig.height=4}

df_delays_fc |>
  dplyr::filter(strain %in% PpcaU_strains) |>
  plot_delay_vs_fc(
    inductiontime = it,
    as_inducer    = aTc,
    legend        = FALSE
  ) +
  ggplot2::annotation_logticks(sides = "l")

ggplot2::ggsave(
  filename = file.path(plot_dir, "1steptimer_foldchange_vs_timedelay_PpcaU_only.pdf"),
  width = 80, height = 50, units = "mm", dpi = 300)

```

---

# Dose-response curves

```{r dose-response, message=FALSE}
for (strain_name in PpcaU_strains) {
  
  plt <- df_av |>
    dplyr::filter(strain == strain_name) |>
    plot_doseresponse()
  
  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("1steptimer_", strain_name, "_doseresponse.pdf")),
    plot     = plt,
    width = 80, height = 45, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Export dose-response data

```{r dose-response-export}

df_doseresponse <- df_av |>
  dplyr::filter(strain %in% strains, time == max(time)) |>
  dplyr::select(strain, antisigma, arabinose, aTc, lum_blank_mean, lum_blank_sd)

readr::write_csv(
  df_doseresponse,
  file.path(plot_dir, "1step_doseresponse_allstrains.csv"))
```

---

# Significance testing

Tests whether threshold induction (aTc or DHBA) significantly changes
maximum fold-change per strain, using a paired t-test on log10(fold-change)
across biological replicates.

```{r significance}
# Ptet strains — aTc threshold
df_atc_sign <- test_threshold_significance(
  .data             = df_blank,
  bg_expr           = bg_expr,
  group_vars        = group_vars,
  strain_group      = strains,
  threshold_inducer = "aTc"
)

# PpcaU strains — DHBA threshold
df_dhba_sign <- test_threshold_significance(
  .data             = df_blank,
  bg_expr           = bg_expr,
  group_vars        = group_vars,
  strain_group      = PpcaU_strains,
  threshold_inducer = "DHBA"
)

# Combine and export
sig_results <- dplyr::bind_rows(
  df_atc_sign  |> dplyr::mutate(inducer = "aTc"),
  df_dhba_sign |> dplyr::mutate(inducer = "DHBA")
)

sig_results

readr::write_csv(
  sig_results,
  file.path(plot_dir, "1step_RW_significance_threshold_vs_no_threshold.csv")
)
```
