---
title: "Anti-sigma and ECF sigma factor toxicity analysis"
author: "Rebecca Wolters"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

# Setup

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(MetBrewer)
library(wellr)       # github: bradyajohnston/wellr
library(broom)
library(wesanderson)
library(patchwork)
library(rstatix)     # for tukey_hsd()
library(here)
```

## Load functions

```{r source}
source(here("ecf_and_as_toxicities", "as_and_ecf_tox_functions_clean.R"))
```

## Experiment parameters

All experiment-level parameters are defined here. Change only this chunk
when analysing a new dataset.

The toxicity experiment measures the effect of expressing sigma (ECF) and
anti-sigma (AS) factors on bacterial growth and luminescence. Key metadata:
- `antisigma` — anti-sigma factor name (e.g. "AS14_1324")
- `tox`       — toxicity level / AS version (e.g. 0 = no AS, 1 = full-length,
                  2.1/2.2 = truncation set 1, 3.1/3.2 = truncation set 2)
- `copies`    — plasmid copy number
- `aTc`       — aTc concentration (ng/ml), drives anti-sigma expression
- `arabinose` — arabinose concentration (%), drives ECF sigma expression

```{r parameters}
# ---- Timing ----------------------------------------------------------------
induction_time <- 2     # time of induction (hours)

# ---- Plate format ----------------------------------------------------------
pc_factor_96well  <- 4.898   # empirical pathlength correction, 100 µl 96-well
pc_factor_384well <- 3.09    # empirical pathlength correction, 50 µl 384-well
pc_factor         <- pc_factor_96well   # change for 384-well experiments

# ---- Blank correction ------------------------------------------------------
# Update average_blank_od from the get_blank_average() output below
average_blank_od <- 0.035 * pc_factor   # 0.035 empirical blank OD for 96-well

# ---- Growth rate window ----------------------------------------------------
# Define the time window used for linear model fitting to log(OD600)
gr_starttime <- 4
gr_endtime   <- 10

# ---- File paths ------------------------------------------------------------
data_files <- here("ecf_and_as_toxicities", "data")
plot_dir   <- here("ecf_and_as_toxicities", "plots")

# ---- Grouping variables ----------------------------------------------------
group_vars    <- c("time", "arabinose", "aTc", "strain", "antisigma", "copies", "tox")
gr_group_vars <- c("strain", "antisigma", "aTc", "arabinose", "rep", "exp", "copies", "tox")
av_group_vars <- c("strain", "aTc", "arabinose", "antisigma", "tox", "copies")

# ---- Strains to exclude from plots -----------------------------------------
# These are controls that are not informative for the toxicity comparison
antisigma_exclude <- c("none", "blank", "Ptet-lux")

# ---- Plot axis breaks -------------------------------------------------------
breaks       <- 10^(-10:10)
minor_breaks <- rep(1:9, 21) * (10^rep(-10:10, each = 9))
```

---

# Data loading

## Raw plate reader data

```{r load-raw}
df_raw <- read_in_plates(data_files, pattern = "tox.csv")
df_raw
```

## Metadata

```{r load-meta}
mdf <- read_in_meta(data_files, pattern = "metainfo.csv")
mdf
```

## Join raw data and metadata

```{r join}
df <- df_raw |>
  dplyr::left_join(mdf, by = c("well" = "Wells", "exp" = "exp")) |>
  na.omit()
df
```

---

# Blanking & averaging

## QC: Check blank readings

```{r check-blank}
check_blank(df,
            blank_variable    = "od600",
            preinduction_time = induction_time,
            variable_to_color = "antisigma")
```

## Get average blank OD

```{r blank-average}
average_blank_od_measured <- get_blank_average(df, preinduction_time = induction_time)
average_blank_od_measured
# Update average_blank_od in the parameters chunk with this value
```

## Apply blank correction

```{r blank-correct}
df_blank <- blank_plate(
  .data            = df,
  average_blank_od = average_blank_od,
  pc_factor        = pc_factor
)
df_blank
```

## QC: Plate overview

```{r plate-overview-od, fig.width=14, fig.height=10}
# Change number_of_experiment to check different plate runs
df_blank |>
  plot_plate(
    variable_to_plot     = od600_blank,
    variable_to_color    = aTc,
    number_of_experiment = 11,
    number_of_colors     = 2
  )
```

```{r plate-overview-lum, fig.width=14, fig.height=10}
df_blank |>
  plot_plate(
    variable_to_plot     = lum_blank,
    variable_to_color    = aTc,
    number_of_experiment = 16,
    number_of_colors     = 2
  ) +
  ggplot2::scale_color_manual(values = c("black", "goldenrod1"))
```

## Average data across replicates

```{r average}
df_av <- df_blank |>
  dplyr::ungroup() |>
  average_plate(group_vars = group_vars)

df_av
```

## Check replicate counts

```{r replicate-counts}
df_av |>
  dplyr::mutate(count = stringr::str_count(replicate, pattern = "[A-Z][0-9]{2}")) |>
  dplyr::summarise(av_count = mean(count), .by = strain)
```

---

# Growth rate and doubling time

## Calculate growth rate

Growth rate is estimated by fitting a linear model to log(OD600) within
the exponential growth window defined by `gr_starttime` and `gr_endtime`.

```{r growthrate}
df_gr <- get_growthrate(
  df_blank,
  group_vars = gr_group_vars,
  starttime  = gr_starttime,
  endtime    = gr_endtime
)

df_gr
```

## Average growth rate across replicates

```{r growthrate-average}
df_gr_av <- df_gr |>
  dplyr::ungroup() |>
  average_gr_data(group_vars = av_group_vars)

# Check wildtype growth rate as a reference
df_gr_av |>
  dplyr::filter(strain == "SV01") |>
  dplyr::summarise(mean_wt_gr = mean(gr_mean, na.rm = TRUE))
```

## Bar plots — doubling time

```{r plot-dt-faceted, fig.width=10, fig.height=6}
df_gr_av |>
  plot_bar_faceted(
    tox_level            = c(3.1, 3.2, 0),
    variable_to_plot     = dt_mean,
    variable_to_plot_sd  = dt_sd,
    y_axis_label         = "dt (h/doubling)"
  )
```

## Bar plots — growth rate (grouped)

```{r plot-gr-grouped, fig.width=10, fig.height=5}
df_gr_av |>
  dplyr::filter(!antisigma %in% antisigma_exclude) |>
  dplyr::mutate(
    interaction_var = factor(
      interaction(aTc, arabinose),
      levels = unique(interaction(aTc, arabinose))[c(1, 2, 3, 4)]
    )
  ) |>
  plot_bar_grouped(
    tox_level           = c(3.1,3.2, 0),  # c(0,1) for FL-P strains, c(0,2.1,2.2) for T1-P and T2-P strains, c(0,3.1,3.2) for T1-C and T2-C strains
    variable_to_plot    = gr_mean,
    variable_to_plot_sd = gr_sd,
    y_axis_label        = "gr (doublings/hour)"
  )

ggplot2::ggsave(
  filename = file.path(plot_dir, "ASECFtox_GR_summary_t1_t2_c.pdf"),
  width = 180, height = 50, units = "mm", scale = 1.5, dpi = 300
)
```

---

# Statistical analysis

## ANOVA + Tukey HSD

Tests whether growth rate differs significantly across inducer combinations
(aTc × arabinose) per strain. Use for comparisons across all four conditions.

```{r anova-tukey}
aov_tukey_results <- run_anova_tukey(
  df_gr,
  antisigma_exclude = antisigma_exclude,
  tox_levels        = c(0, 1)
)

aov_tukey_results

readr::write_csv(
  aov_tukey_results,
  file.path(plot_dir, "anova_tukey_results_fl_p.csv")
)
```

## Paired t-tests — aTc effect per antisigma

Tests whether AS induction (aTc) significantly reduces growth rate,
run separately per antisigma factor.

```{r t-tests}
t_test_results <- run_t_tests(
  df_gr_av,
  antisigma_exclude = antisigma_exclude,
  tox_levels        = c(0, 3.1, 3.2),
  atc_low           = 0,
  atc_high          = 100
)

t_test_results

readr::write_csv(
  t_test_results,
  file.path(plot_dir, "t_test_results_atc_effect.csv")
)
```

---

# Growth curves

## Kinetic OD — per antisigma, filtered by toxicity level

```{r growthcurves-faceted, fig.width=10, fig.height=8}
df_av |>
  dplyr::filter(!antisigma %in% antisigma_exclude) |>
  plot_growthcurves(tox_level = c(3.2,3.1, 0))

ggplot2::ggsave(
  filename = file.path(plot_dir, "ASECFtox_growthcurve_t1_t2_c.png"),
  width = 200, height = 200, units = "mm", scale = 1.5, dpi = 300
)

```

## Kinetic OD — all strains overlaid

```{r growthcurves-allinone, fig.width=8, fig.height=5}
df_av |>
  dplyr::filter(strain != "blank") |>
  plot_growthcurves_allinone() +
  ggplot2::facet_grid(cols = ggplot2::vars(interaction(aTc, arabinose)))
```

## Combination plot — example comparing truncated vs full-length AS

```{r growthcurves-combo, fig.width=14, fig.height=6}
# Update these to match the antisigma names in your dataset
untruncated_AS <- "AS26_4464"
truncated_AS   <- c("AS26_t1", "AS26_t2")

plot_1 <- df_av |>
  dplyr::filter(antisigma == untruncated_AS) |>
  plot_growthcurves(tox_level = 1)

plot_2 <- df_av |>
  dplyr::filter(antisigma %in% truncated_AS) |>
  plot_growthcurves(tox_level = c(2.1, 2.2)) +
  ggplot2::facet_wrap(facets = "antisigma", nrow = 2)

plot_3 <- df_av |>
  dplyr::filter(antisigma %in% truncated_AS) |>
  plot_growthcurves(tox_level = c(3.1, 3.2)) +
  ggplot2::facet_wrap(facets = "antisigma", nrow = 2)

plot_1 + plot_2 + plot_3 +
  patchwork::plot_annotation(tag_levels = "A")
```

---

# Luminescence

## Endpoint luminescence bar plot

Shows luminescence at the induction time point across antisigma factors
and arabinose concentrations.

```{r endpoint-lum, fig.width=10, fig.height=6}
df_av |>
  dplyr::filter(
    !antisigma %in% antisigma_exclude,
    strain     != "SV01"
  ) |>
  plot_endpoint_lum(tox_level = c(3.1, 3.2, 0))

ggplot2::ggsave(
  filename = file.path(plot_dir, "ASECFtox_EndLUM.pdf"),
  width = 35, height = 15, units = "cm", dpi = 300
)
```

## Kinetic luminescence

```{r kinetic-lum, fig.width=10, fig.height=8}
# Get background luminescence from blank strain for reference line
background_lum <- df_av |>
  dplyr::filter(strain == "blank") |>
  dplyr::summarise(mean_lum = mean(lum_blank_mean, na.rm = TRUE)) |>
  dplyr::pull(mean_lum)

background_lum

df_av |>
  plot_kinetic_lum(tox_level = c(2.1, 2.2)) +
  ggplot2::geom_hline(
    yintercept = background_lum,
    linetype   = "dashed",
    colour     = "tomato"
  )

#ggsave(plot_dir, filename = "ASECFtox_Growthcurve_chrint_truncated.png", width= 20, height = 20, unit = "cm", dpi = 300)
```

## Combination plot — luminescence kinetics

```{r kinetic-lum-combo, fig.width=14, fig.height=6}
plot_lum_1 <- df_av |>
  dplyr::filter(antisigma == untruncated_AS) |>
  plot_kinetic_lum(tox_level = 1)

plot_lum_2 <- df_av |>
  dplyr::filter(antisigma %in% truncated_AS) |>
  plot_kinetic_lum(tox_level = c(2.1, 2.2))

plot_lum_3 <- df_av |>
  dplyr::filter(antisigma %in% truncated_AS) |>
  plot_kinetic_lum(tox_level = c(3.1, 3.2))

plot_lum_1 + plot_lum_2 + plot_lum_3 +
  patchwork::plot_annotation(tag_levels = "A")
```

## Luminescence fold-repression

Calculates fold-repression as luminescence without aTc / luminescence with aTc,
at maximum arabinose and the final time point.

```{r fold-repression, fig.width=10, fig.height=5}
df_fr <- get_fold_repression(df_av, atc_uninduced = 0, atc_induced = 100)

df_fr |>
  dplyr::filter(tox %in% c(3.1, 3.2, 0)) |>
  ggplot2::ggplot(ggplot2::aes(
    x    = reorder(as.factor(antisigma), dplyr::desc(fc_mean)),
    y    = fc_mean,
    fill = as.factor(tox)
  )) +
  ggplot2::geom_col(
    stat     = "identity",
    width    = 0.5,
    position = ggplot2::position_dodge(width = 0.5),
    colour   = "black"
  ) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      ymin = fc_mean - fc_sd,
      ymax = fc_mean + fc_sd
    ),
    width    = 0.2,
    position = ggplot2::position_dodge(width = 0.5),
    colour   = "black"
  ) +
  ggplot2::scale_fill_manual(values = c("grey50", "black")) +
  ggplot2::labs(
    y    = "fold-repression",
    fill = "AS version"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x      = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 0.5),
    legend.position  = "right"
  )

ggplot2::ggsave(
  filename = file.path(plot_dir, "ASECFtox_LUM_foldrepression.pdf"),
  width = 200, height = 75, units = "mm", scale = 1.5, dpi = 300
)
```

---

# Toxicity heatmaps

## Toxicity relative to WT — summary heatmap

```{r tox-heatmap, fig.width=12, fig.height=6}

percent_diff <- function(wt_od, gmo_od) {
    0-(((wt_od-gmo_od)/wt_od)*100)
 }
 

df_tox_dif <- df_av|>
  dplyr::filter(time == max(time)) |>
  dplyr::mutate(as_origin = stringr::str_extract(antisigma, "([A-Z]{2}\\d{2})"),
                toxicity_blank = (percent_diff(wt_od = od600_blank_mean[strain == "SV01"],
                                               gmo_od =od600_blank_mean))) |>
  dplyr::filter(strain !="SV01") |> 
  drop_na(as_origin)|> 
  average_tox_data(group_vars = c("aTc","arabinose", "strain", "antisigma","copies", "tox", "as_origin")) |> 
  dplyr::mutate(
    interaction_var = factor(interaction(aTc, arabinose), 
                             levels = unique(interaction(aTc, arabinose))[c(2,4,1,3)])
    )

df_tox_dif|>  
  ggplot2::ggplot(aes(x = interaction_var, y = as.factor(tox), fill = toxicity_blank_mean))+
  ggplot2::geom_tile()+
  ggplot2::scale_fill_gradient2( low="brown2" ,mid = "white" , high = "cornflowerblue", 
                                 midpoint = 0
                                )+
  ggplot2::scale_x_discrete(expand = c(0,0))+
  ggplot2::scale_y_discrete(expand = c(0,0))+
  ggplot2::facet_wrap(facets = vars(as_origin), 
                        nrow = 1)  +
  ggplot2::labs(fill = "Toxcicty (% OD difference compared to WT)")+
   ggplot2::theme_bw()+
  ggplot2::theme(axis.text.x = element_text(angle = 45,
                                            vjust = 0.5,
                                              hjust=0.5, 
                                             size = 10))

ggplot2::ggsave(
  filename = file.path(plot_dir, "ASECF_toxicity_WT_OD_summarised_heatmap.pdf"),
  width = 300, height = 50, units = "mm", scale = 2, dpi = 300
)

```
