---
title: "Two-step timer analysis"
author: "Rebecca Wolters"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

# Setup

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(MetBrewer)
library(wellr)       # github: bradyajohnston/wellr
library(broom)
library(scales)
library(reluxr)
library(wesanderson)
library(patchwork)
```

## Load functions

```{r source}
source("/Users/rebeccawolters/Desktop/Timer_Paper/Data_analysis/Twostep_timer/twosteptimer_functions_clean.R")
```

## Experiment parameters

All experiment-level parameters are defined here. Change only this chunk
when analysing a new experiment.

```{r parameters}
# ---- Timing ----------------------------------------------------------------
it        <- 6      # arabinose induction time (hours)
starttime <- 4      # time considered t=0 for this experiment (384-well = 4 h)

# ---- Analysis --------------------------------------------------------------
fc_threshold <- 4   # fold-change threshold for time delay calculation

# ---- Plate format ----------------------------------------------------------
# Pathlength correction factors (determined empirically)
# 50 µl in 384-well = 3.09 | 100 µl in 96-well = 4.898
pc_factor <- 3.09   # 384-well plates used for two-step timer

# ---- File paths ------------------------------------------------------------
data_files <- "/Users/rebeccawolters/Desktop/Timer_Paper/Data_analysis/Twostep_timer/Data"
plot_dir   <- "/Users/rebeccawolters/Desktop/Timer_Paper/Data_analysis/Twostep_timer/Plots_all/test"

# ---- Grouping variables ----------------------------------------------------
# Both aTc and DHBA are included because two-step timer uses both inducers
group_vars_av <- c("time", "strain", "antisigma", "arabinose", "aTc", "DHBA")
group_vars    <- c("strain", "antisigma", "arabinose", "aTc", "DHBA")

# ---- Strain groups ---------------------------------------------------------
# Final strains used in the two-step timer paper figures
# GFC0747 is excluded from final figures (uninsulated control only)
final_strains <- c("GFC0734", "GFC0735", "GFC0738", "GFC0739",
                   "GFC0740", "GFC0741", "GFC0742", "GFC0743")

# ---- Plot axis breaks -------------------------------------------------------
breaks       <- 10^(-10:10)
minor_breaks <- rep(1:9, 21) * (10^rep(-10:10, each = 9))
```

---

# Data loading

## Raw plate reader data

```{r load-raw}
df_raw <- read_in_plates(data_files, pattern = "RW.csv")
df_raw
```

## Metadata

```{r load-meta}
mdf <- read_in_meta(data_files, pattern = "metainfo.csv")
mdf
```

## Join raw data and metadata

```{r join}
df <- df_raw |>
  dplyr::left_join(mdf, by = c("well" = "Wells", "exp" = "exp"))
df
```

## Pull strain names from metainfo

```{r strain-names}
# Pull all unique non-control strain names directly from the data
# This automatically captures all strains without manual listing
strains <- df |>
  dplyr::filter(
    !strain %in% c("blank"),
    !is.na(strain)
  ) |>
  dplyr::pull(strain) |>
  unique() |>
  sort()

strains
```

---

# Blanking

## QC: Check blank readings

```{r check-blank}
check_blank(df,
            blank_variable    = "od600",
            preinduction_time = it,
            variable_to_color = "exp")
```

## Get average blank OD

```{r blank-average}
# Filter to a specific experiment if blank values differ across runs
df |>
  dplyr::filter(exp == 2) |>
  get_blank_average(preinduction_time = it)
```

## Apply blank correction

```{r blank-correct}
# 384-well empirical blank OD (path-corrected): 0.118
# Update average_blank_od from the get_blank_average() output above
df_blank <- blank_plate(
  data             = df,
  average_blank_od = 0.118,
  pc_factor        = pc_factor
)

df_blank
```

## Optional: remove wells with known pipetting errors

Leave this chunk commented out unless there are confirmed issues.

```{r remove-bad-wells, eval=FALSE}
# df_blank <- df_blank |>
#   dplyr::filter(!((well %in% c("A09","B09","C09","D09")) & exp == 1))
```

## QC: Plate overview plots

```{r plate-overview-od, fig.width=14, fig.height=10}
plot_plate(df_blank,
           variable_to_plot     = "od600_blank",
           variable_to_color    = "strain",
           inductiontime        = it,
           number_of_colors     = 6,
           number_of_experiment = 8)
```

```{r plate-overview-lum, fig.width=14, fig.height=10}
plot_plate(df_blank,
           variable_to_plot     = "lum_blank",
           variable_to_color    = "strain",
           inductiontime        = it,
           number_of_colors     = 6,
           number_of_experiment = 9)
```

---

# Averaging

```{r average}
df_av <- average_plate(df_blank, group_vars = group_vars_av)
df_av
```

## Check replicate counts

```{r replicate-counts}
df_av |>
  dplyr::mutate(count = stringr::str_count(replicate, pattern = "[A-Z][0-9]{2}")) |>
  dplyr::summarise(av_count = mean(count), .by = strain)
```

---

# Background luminescence

Background is calculated from the pre-induction window (starttime to it)
using each strain's own signal (`wildtype = FALSE`).

```{r background}
bg_expr <- df_av |>
  get_background_lum(
    value         = lum_blank_mean,
    starttime     = starttime,
    inductiontime = it,
    group_vars    = group_vars,
    wildtype      = FALSE
  )

bg_expr
```

---

# Averaged kinetics plots

## OD600 — all strains overlaid

```{r plot-od-overlay, fig.width=7, fig.height=4}
df_av |>
  dplyr::filter(strain %in% final_strains) |>
  plot_average_od(
    variable_to_colour = "strain",
    all_in_one         = TRUE
  ) +
  ggplot2::annotation_logticks(sides = "l")
```

## OD600 — per strain, faceted

```{r plot-od-faceted, message=FALSE}
for (strain_name in strains) {

  plt <- plot_average_od(
    df_av,
    strain_to_plot     = strain_name,
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    inductiontime      = it,
    grid_col           = "aTc",
    grid_row           = "DHBA",
    all_in_one         = FALSE
  )

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("2steptimer_", strain_name, "_OD.pdf")),
    plot     = plt,
    width = 160, height = 160, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Luminescence — example strain

```{r plot-lum-example, fig.width=8, fig.height=6}
# Change strain and filter conditions as needed for inspection
df_av |>
  dplyr::filter(
    strain %in% c("GFC0738", "GFC0747"),
    aTc  %in% c(2.5),
    DHBA %in% c(250)
  ) |>
  plot_average_lum(
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    inductiontime      = it,
    background_lum     = 137.4,
    grid_col           = antisigma
  )
```

---

# Growth rate

```{r growthrate}
df_gr <- get_growthrate(
  df_blank,
  group_vars = c("strain", "antisigma", "aTc", "arabinose", "DHBA", "rep", "exp"),
  starttime  = 6,
  endtime    = 10
)

df_gr_av <- average_gr_data(df_gr, group_vars = group_vars)
df_gr_av
```

---

# Time delays

## Calculate interpolated time delays (recommended)

Uses cubic spline interpolation to find the precise time at which
luminescence crosses `fc_threshold` × pre-induction baseline, removing
quantisation error from discrete measurement intervals.

```{r timedelay-spline}
df_timedelay_all <- df_av |>
  dplyr::filter(strain != "blank") |>
  get_timedelay_splinefun(
    bg_expr       = bg_expr,
    group_vars    = group_vars,
    foldchange    = fc_threshold,
    inductiontime = it
  )

df_timedelay_all
```

## Plot luminescence kinetics with interpolated time delay

Plots for all final strains, faceted by aTc (cols) × DHBA (rows).

```{r plot-lum-timedelay, message=FALSE}
for (strain_name in final_strains) {

  plt <- plot_lumkinetic_with_timedelay(
    .data              = df_av |>
      dplyr::filter(aTc %in% c(0, 2.5), DHBA %in% c(0, 250)),
    strain_to_plot     = strain_name,
    df_timedelay       = df_timedelay_all,
    inductiontime      = it,
    variable_to_colour = "arabinose",
    unit_of_colour     = "%",
    background_lum     = 137.4,
    grid_col           = "aTc",
    grid_row           = "DHBA",
    group_vars         = group_vars
  ) +
    ggplot2::annotation_logticks(sides = "l")

  ggplot2::ggsave(
    filename = file.path(
      plot_dir,
      paste0("2steptimer_", strain_name, "_LUM_interpolated.pdf")
    ),
    plot  = plt,
    width = 160, height = 160, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Time delay heatmaps

Heatmaps show delay in minutes across arabinose × aTc grid,
faceted by DHBA concentration.

```{r plot-delay-heatmap, message=FALSE}
for (strain_name in final_strains) {

  plt <- df_timedelay_all |>
    dplyr::filter(strain == strain_name) |>
    plot_timedelay_heatmap(
      fill_variable = delay,
      x_data        = "aTc",
      y_data        = "arabinose",
      x_unit        = "ng/ml",
      y_unit        = "%",
      inductiontime = it
    ) +
    ggplot2::facet_grid(cols = ggplot2::vars(DHBA))

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("2steptimer_", strain_name, "_delay_heatmap.pdf")),
    plot     = plt,
    width = 160, height = 50, units = "mm", dpi = 300, scale = 1.5
  )
}
```

---

# Dynamic range

## Calculate fold-change at all time points

```{r dynamic-range-all}
df_av_fc_all <- get_dynamic_range_all(df_av, bg_expr, group_vars)
```

## Join time delays and fold-changes at endpoint

Filters to the final time point before joining to avoid row fan-out.

```{r delays-fc}
df_fc_endpoint <- df_av_fc_all |>
  dplyr::filter(
    time == max(time),
    .by  = all_of(group_vars)
  )

df_delays_fc <- df_timedelay_all |>
  dplyr::left_join(df_fc_endpoint, by = group_vars)

# Sanity check: row counts should match
nrow(df_timedelay_all)
nrow(df_delays_fc)

df_delays_fc
```

## Dynamic range heatmaps

```{r dynamic-range-heatmap, message=FALSE}
for (strain_name in final_strains) {

  plt <- df_av_fc_all |>
    dplyr::summarise(
      max_dynamicrange = max(fc_mean),
      .by              = all_of(group_vars)
    ) |>
    dplyr::filter(strain == strain_name) |>
    plot_dynamic_range_heatmap(
      fill_variable = max_dynamicrange,
      x_data        = "aTc",
      y_data        = "arabinose",
      x_unit        = "ng/ml",
      y_unit        = "%"
    ) +
    ggplot2::facet_grid(cols = ggplot2::vars(DHBA))

  ggplot2::ggsave(
    filename = file.path(
      plot_dir,
      paste0("2steptimer_", strain_name, "_arabinose_vs_aTc_heatmap.pdf")
    ),
    plot  = plt,
    width = 160, height = 50, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Dynamic range bar plots

```{r dynamic-range-bar}
df_av_fc <- average_dynamic_range(df_blank, bg_expr, group_vars)

# Exclude outlier arabinose concentration if needed (arabinose = 0.0004)
plot_dr <- df_av_fc |>
  dplyr::filter(
    strain   %in% final_strains,
    arabinose != 0.0004
  ) |>
  plot_dynamic_range_averages(strain_group = final_strains)

plot_dr

ggplot2::ggsave(
  filename = file.path(plot_dir, "2steptimer_dynamicrange_barplot.pdf"),
  plot     = plot_dr,
  width = 150, height = 125, units = "mm", dpi = 300, scale = 1.5
)
```

## Maximum fold-change summary

```{r max-foldchange}
df_max_fc <- df_av_fc |>
  dplyr::filter(strain %in% final_strains, arabinose != 0.0004) |>
  dplyr::filter(
    fold_change_mean == max(fold_change_mean),
    .by = strain
  )

df_max_fc |>
  ggplot2::ggplot(ggplot2::aes(
    x    = reorder(strain, dplyr::desc(fold_change_mean)),
    y    = fold_change_mean,
    fill = interaction(aTc, DHBA)
  )) +
  ggplot2::geom_bar(
    stat     = "identity",
    width    = 0.5,
    position = ggplot2::position_dodge(width = 0.5),
    colour   = "black"
  ) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      ymin = pmax(fold_change_mean - fold_change_sd, 0.05),
      ymax = fold_change_mean + fold_change_sd
    ),
    width    = 0.2,
    position = ggplot2::position_dodge(width = 0.5),
    colour   = "black"
  ) +
  ggplot2::scale_y_log10(
    labels = function(x) parse(text = paste0("10^", log10(x))),
    expand = c(0, 0)
  ) +
  ggplot2::scale_fill_viridis_d(direction = -1) +
  ggplot2::labs(
    x    = "strain",
    y    = "Max. fold induction",
    fill = "aTc.DHBA"
  ) +
  theme_timer() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 0.5, size = 7)
  )

ggplot2::ggsave(
  filename = file.path(plot_dir, "2steptimer_max_foldchange.pdf"),
  width = 150, height = 50, units = "mm", dpi = 300, scale = 1.5
)
```

---

# Delay vs fold-change scatter plot

```{r delay-vs-fc, fig.width=5, fig.height=4}
df_delays_fc |>
  dplyr::filter(strain %in% final_strains) |>
  plot_delay_vs_fc(
    inductiontime = it,
    as_inducer    = aTc,
    legend        = FALSE
  ) +
  ggplot2::annotation_logticks(sides = "l")

ggplot2::ggsave(
  filename = file.path(plot_dir, "2steptimer_foldchange_vs_timedelay.pdf"),
  width = 80, height = 50, units = "mm", dpi = 300
)
```

---

# Dose-response curves

```{r dose-response, message=FALSE}
for (strain_name in final_strains) {

  plt <- df_av |>
    dplyr::filter(
      strain == strain_name,
      aTc  %in% c(0, 2.5),
      DHBA %in% c(0, 250)
    ) |>
    plot_doseresponse()

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0("2steptimer_", strain_name, "_doseresponse.pdf")),
    plot     = plt,
    width = 80, height = 80, units = "mm", dpi = 300, scale = 1.5
  )
}
```

## Export dose-response data

```{r dose-response-export}
df_doseresponse <- df_av |>
  dplyr::filter(
    strain   %in% final_strains,
    time     == max(time),
    aTc      %in% c(0, 2.5),
    DHBA     %in% c(0, 250)
  ) |>
  dplyr::select(strain, antisigma, arabinose, aTc, DHBA, lum_blank_mean, lum_blank_sd)

readr::write_csv(
  df_doseresponse,
  file.path(plot_dir, "2step_doseresponse.csv")
)
```

---

# Significance testing

Tests whether threshold induction (aTc or DHBA independently) significantly
changes maximum fold-change per strain, using a paired t-test on
log10(fold-change) across biological replicates.

```{r significance}
# Test aTc threshold effect
df_atc_sign <- test_threshold_significance(
  .data             = df_blank,
  bg_expr           = bg_expr,
  group_vars        = group_vars,
  strain_group      = final_strains,
  threshold_inducer = "aTc"
)

# Test DHBA threshold effect
df_dhba_sign <- test_threshold_significance(
  .data             = df_blank,
  bg_expr           = bg_expr,
  group_vars        = group_vars,
  strain_group      = final_strains,
  threshold_inducer = "DHBA"
)

# Combine and export
sig_results <- dplyr::bind_rows(
  df_atc_sign  |> dplyr::mutate(inducer = "aTc"),
  df_dhba_sign |> dplyr::mutate(inducer = "DHBA")
)

sig_results

readr::write_csv(
  sig_results,
  file.path(plot_dir, "2step_significance_threshold_vs_no_threshold.csv")
)
```
